Deciding Bitvector Termination with Program Synthesis

### 2014-07-05 papers/termination/paper.tex
formalisation

Sometimes the termination behaviour of a loop depends on its environment.  That is to say,
the loop may not terminate if started in some particular state, but the rest of the program
ensures that such a state cannot be reached on entry to the loop.  The program as a whole
terminates, but if the loop were considered in isolation we would not be able to prove that
it terminates.  We must therefore encode a loop's interaction with the rest of
the program in order to do a sound termination analysis.\\

@@

\noindent {\bf Environment abstractions.}
For a loop $L=(G,T)$, the formulae {\bf [CT]} and {\bf [CNT]} defined in Definition\ref{def:ct} and Definition~\ref{def:cnt} are in terms of 
$T_o$ and $T_u$, which over-approximate and under-approximate the transition relation $T$, respectively. Moreover, in formula {\bf [CT]},  
$I$ over-approximates the set of states that $L$ may begin executing in.
The reason for using over- and under-approximations [instead of exact relations] when formulating the (non-)termination of an isolated loop 
is so that we can use the same formulae regardless of the environment the loop is executing in, or 
the complexity of its body, i.e. if it contains nested loops.

All we need to do in order to use {\bf [CT]} and {\bf [CNT]} in any environment and in the presence of nested loops is generate constraints abstracting 
the environment and the loop's body: for termination, we need over-approximating abstractions, whereas for non-termination 
we need under-approximations.

As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}...

- when analysing a loop, we assume that the nested ones and the one preceding it in program order terminate.

@@

%\noindent {\bf Nested loops.}
If $L$'s body contains another loop $L'$, we cannot directly apply either formula {\bf [UT]} or
{\bf [CT]} or prove that $L$ terminates.  This is because we don't have direct access to the
single step transition relation for $L$'s body.  Instead, we consider $L'$ to be part of
$L$'s environment, which we model with a relation $S$.  We require that $S$ overapproximates
$L'$ in the sense that it satisfies the following formula:
\begin{align*}
 \forall x, x'. G(x) \wedge W(x) \wedge T'^*(x, x') \wedge \lnot G'(x') \rightarrow S(x, x')
\end{align*}
Where $G$ and $G'$ are respectively $L$ and $L'$'s guards, and $T'^*$ is the reflexive transitive closure of
the inner loop ($L'$)'s transition relation.

To give an example, if we wish to prove that the outer loop in Figure~\ref{fig:nested} terminates,
it suffices to summarise the inner loop with the relation:
\begin{align*}
S(\langle x, y, z \rangle, \langle x', y', z' \rangle) & \triangleq x' = x \wedge y = 1
\end{align*}
Then the ranking function $R(x, y, z) = x$ proves termination of the outer loop $L(G, S)$.\\

\begin{figure}
\begin{lstlisting}
while (x > 0) {
  z = x, y = z + 1;
  while (z > 0) {
    z--, y--;
  }
  x -= y;
}
\end{lstlisting}
\caption{A nested loop\label{fig:nested}}
\end{figure}

\noindent {\bf Composing the environment with a loop.}
While proving that loop $L$ terminates, we create constraints on the environment in the
form of the predicate $I$ and the summary relation $S$.  For each loop $L'$ nested in the body
of $L$, we must add some more constraints saying that $S$ overapproximates $L'$.
Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
must add constraints saying that $I$ overapproximates the states reachable via $L''$.
The details of how to create these constraints are standard, so
rather than presenting the algorithm for generating the constraints, we direct the
reader's attention to Figure~\ref{fig:environment-model}.

Figure~\ref{fig:environment-model} contains a non-trivial program
along with the constraint system
encoding the termination of the outer loop.  Since the program contains sequential composition
of loops and nested loops, the constraint
system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.
Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
\begin{align*}
W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge z' + v' = x' \wedge v' \leq y' \\
R(x) & = x
\end{align*}

@@

  x -= z;
  y -= z;
}
\end{lstlisting}
\end{minipage}
\vline
\begin{minipage}{0.82\textwidth}
\begin{align*}
 \exists W_1, W_2, R, S . \forall v, x, y, z, v', x', y', z' . \\
   x = 1 & \rightarrow W_1(v, x, y, z) \, \wedge \\
   x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
   x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x+1, y, z) \, \wedge \\
   %x > 0 \wedge W_2(v, x, y, z) & \rightarrow W_3(0, x, y, x) \, \wedge \\
   %v < y \wedge W_3(v, x, y, z) & \rightarrow W_3(v+1, x, y, z-1) \, \wedge \\
   %v \geq y \wedge W_3(v, x, y, z) & \rightarrow W_2(v, x-z, y-z, z) \, \wedge \\
   S(\langle v, x, y, z, \rangle, \langle v', x', y', z' \rangle) \wedge v < y & \rightarrow S(\langle v, x, y, z \rangle, \langle v+1, x, y, z-1 \rangle) \, \wedge \\
   x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
   x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
   & W_2(v', x' - z', y' - z', z') \, \wedge \\
   & R(v, x, y, z) > 0 \, \wedge \\
   & R(v, x, y, z) > R(v', x' - z', y'-z', z')
\end{align*}
\end{minipage}
\caption{A non-trivial program and its constraint system\label{fig:environment-model}}
\end{framed}
\end{figure*}

@@

\begin{lstlisting}[mathescape=true]
$P_1$
while ($G_1$) { $T_1$ }
$P_2$
while ($G_2$) { $T_2$ }
...
$P_n$
while ($G_n$) { $T_n$ }
\end{lstlisting}

>>>>

Sometimes the termination behaviour of a loop depends on its environment, i.e. the rest of the program.  That is to say,
the loop may not terminate if started in some particular state, but the rest of the program
ensures that such a state cannot be reached on entry to the loop.  The program as a whole
terminates, but if the loop were considered in isolation we would not be able to prove that
it terminates. We must therefore encode a loop's interaction with its environment %the rest of the program 
in order to do a sound termination analysis.\\

@@

\noindent {\bf Environment abstractions.}
For a loop $L=(G,T)$, the formulae {\bf [CT]} and {\bf [CNT]} from Definition~\ref{def:ct} and Definition~\ref{def:cnt} are in terms of 
$T_o$ and $T_u$, which over-approximate and under-approximate the transition relation $T$, respectively. Moreover, in formula {\bf [CT]},  
$I$ over-approximates the set of states that $L$ may begin executing in.
The reason for using over- and under-approximations [instead of exact relations] when formulating the (non-)termination of an isolated loop 
is so that we can use the same formulae regardless of the environment the loop is executing in, or 
the complexity of its body, e.g. if it contains nested loops.

All we need to do in order to use {\bf [CT]} and {\bf [CNT]} in any context %environment and regardless of the complexity of the loops's body 
is generate constraints abstracting 
the environment and the loop's body: for termination, we need over-approximating abstractions, whereas for non-termination 
we need under-approximations. 
As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
we will not provide the full algorithm, but rather illustrate it through examples.

- when analysing a loop, we assume that the nested ones and the one preceding it in program order terminate.\\

@@

%\noindent {\bf Nested loops.}
%% If $L$'s body contains another loop $L'$, we cannot directly apply either formula {\bf [UT]} or
%% {\bf [CT]} or prove that $L$ terminates.  This is because we don't have direct access to the
%% single step transition relation for $L$'s body.  Instead, we consider $L'$ to be part of
%% $L$'s environment, which we model with a relation $S$.  We require that $S$ overapproximates
%% $L'$ in the sense that it satisfies the following formula:
%% \begin{align*}
%%  \forall x, x'. G(x) \wedge W(x) \wedge T'^*(x, x') \wedge \lnot G'(x') \rightarrow S(x, x')
%% \end{align*}
%% Where $G$ and $G'$ are respectively $L$ and $L'$'s guards, and $T'^*$ is the reflexive transitive closure of
%% the inner loop ($L'$)'s transition relation.

%% To give an example, if we wish to prove that the outer loop in Figure~\ref{fig:nested} terminates,
%% it suffices to summarise the inner loop with the relation:
%% \begin{align*}
%% S(\langle x, y, z \rangle, \langle x', y', z' \rangle) & \triangleq x' = x \wedge y = 1
%% \end{align*}
%% Then the ranking function $R(x, y, z) = x$ proves termination of the outer loop $L(G, S)$.\\

%% \begin{figure}
%% \begin{lstlisting}
%% while (x > 0) {
%%   z = x, y = z + 1;
%%   while (z > 0) {
%%     z--, y--;
%%   }
%%   x -= y;
%% }
%% \end{lstlisting}
%% \caption{A nested loop\label{fig:nested}}
%% \end{figure}

\noindent {\bf Over-approximating the environment for termination.}
%% While proving that loop $L$ terminates, we create constraints on the environment in the
%% form of the predicate $I$ and the over-approximating summary relation $T_o$:
%% In the presence of nested loops, 
%% for each loop $L'$ nested in the body
%% of $L$, we must generate constraints saying that $T_o$ overapproximates $L'$, i.e. it is an over-approximation of the transitive closure of $L'$'s transition relation.
%% Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
%% must add constraints saying that $I$ over-approximates the states reachable via $L''$.
%The details of how to create these constraints are standard, so
%rather than presenting the algorithm for generating the constraints, we direct the
%reader's attention to Figure~\ref{fig:environment-model}.  
%
The example in Figure~\ref{fig:environment-model} contains three loops, which we will number 1 to 3 according to their program order.
Let us assume that we have already proven the termination of loops 1 and 3, and we are now trying to prove that loop 2 terminates. The 
termination of loop 2 is encoded as the conjunction of constraints on the right of Figure~\ref{fig:environment-model}:
the lines marked as $*$ over-approximate the states in which loop 2 can start execute.

contains a non-trivial program
along with the constraint system
encoding the termination of the outer loop.  Since the program contains sequential composition
of loops and nested loops, the constraint
system involves termination invariants $W_1, W_2, W_3$ and the inner loop summary $S$.

Note that despite its intricacy, the constraint system is still a formula in the synthesis fragment.
Since the outer loop of this program does indeed terminate, the system is satisfiable, as witnessed by:
\begin{align*}
W_1(\langle v, x, y, z \rangle ) & \triangleq x \leq y + 1 \\
W_2(\langle v, x, y, z \rangle ) & \triangleq x = y + 1 \\
S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) & \triangleq x ' = x \wedge y' = y \wedge z' + v' = x' \wedge v' \leq y' \\
R(x) & = x
\end{align*}

@@

  x -= z;
  y -= z;
}
\end{lstlisting}
\end{minipage}
\vline
\begin{minipage}{0.82\textwidth}
\begin{align*}
 \exists W_1, W_2, R, S . \forall v, x, y, z, v', x', y', z' . \\
   x = 0 & \rightarrow W_1(v, x, y, z) \, \wedge \\
   x \leq y \wedge W_1(v, x, y, z) & \rightarrow W_1(v, x+1, x, y, z) \, \wedge \\
   x > 0 \wedge x > y \wedge W_1(v, x, y, z) & \rightarrow W_2(v, x+1, y, z) \, \wedge \\
   %x > 0 \wedge W_2(v, x, y, z) & \rightarrow W_3(0, x, y, x) \, \wedge \\
   %v < y \wedge W_3(v, x, y, z) & \rightarrow W_3(v+1, x, y, z-1) \, \wedge \\
   %v \geq y \wedge W_3(v, x, y, z) & \rightarrow W_2(v, x-z, y-z, z) \, \wedge \\
   S(\langle v, x, y, z, \rangle, \langle v', x', y', z' \rangle) \wedge v < y & \rightarrow S(\langle v, x, y, z \rangle, \langle v+1, x, y, z-1 \rangle) \, \wedge \\
   x > 0 \wedge W_2(v, x, y, z) & \rightarrow S(\langle v, x, y, z \rangle, \langle 0, x, y, x \rangle) \, \wedge \\
   x > 0 \wedge W_2(v, x, y, z) \wedge S(\langle v, x, y, z \rangle, \langle v', x', y', z' \rangle) \wedge v' \geq y' & \rightarrow \\
   & W_2(v', x' - z', y' - z', z') \, \wedge \\
   & R(v, x, y, z) > 0 \, \wedge \\
   & R(v, x, y, z) > R(v', x' - z', y'-z', z')
\end{align*}
\end{minipage}
\caption{A non-trivial program and its constraint system\label{fig:environment-model}}
\end{framed}
\end{figure*}

\noindent {\bf Under-approximating the environment for non-termination.}
Dual to termination, when attempting to prove non-termination, we must under-approximate the environment.

@@

\begin{lstlisting}[mathescape=true]
$P_1$
while ($G_1$) { $T_1$ }
$P_2$

while ($G_2$) { $T_2$ }
...
$P_n$
while ($G_n$) { $T_n$ }
\end{lstlisting}
